Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-04-22 thesis/draft.tex
Abstract star

\newpage
\section{Abstract}
Simple dynamical models based on Ordinary Differential Equations (ODE) are widely utilized to forecast the spread of infectious diseases and to analyze other biological systems. A limitation of this framework is its implicit assumption that residence time in any given state, such as the infectious state, follows an exponentially distributed pattern. This assumption is often not realistic. A common solution is the application of the Linear Chain Trick (LCT), wherein a single stage is subdivided into multiple substages, each of which follows an identical exponential distribution, thereby yielding a more realistic Erlang distribution for the duration. However, the LCT poses challenges by restricting parametric flexibility, particularly resulting in less flexible dwell-time distributions, and it often necessitates comparisons between models with varying structures (i.e. different numbers of substages). Here, we propose a Geometric Chain Trick (GCT) employing a fixed number of substages with transition rates forming a geometric sequence. Through mathematical analysis, simulation, and model fitting, we demonstrate that our approach retains the principal benefits of the LCT while providing increased flexibility and facilitating easier data fitting. Compared to the recently introduced Generalized Linear Chain Trick (GLCT), which incorporates higher-dimensional auxiliary stages to model delays in transitions between main compartments, our method offers a more direct means of achieving flexibility. Although this study focuses on the application of the LCT and GCT in the epidemic SIR (susceptible-infectious-recovered) model, the potential applications of our approach span a broad array of dynamical models, including those for infectious diseases and other systems (e.g. population dynamics, biochemical reaction networks), offering advantages in computational efficiency, ease of parameter estimation, and the flexibility of time distributions.

>>>>

\newpage
\section*{Abstract}
Simple dynamical models based on Ordinary Differential Equations (ODE) are widely utilized to forecast the spread of infectious diseases and to analyze other biological systems. A limitation of this framework is its implicit assumption that residence time in any given state, such as the infectious state, follows an exponentially distributed pattern. This assumption is often not realistic. A common solution is the application of the Linear Chain Trick (LCT), wherein a single stage is subdivided into multiple substages, each of which follows an identical exponential distribution, thereby yielding a more realistic Erlang distribution for the duration. However, the LCT poses challenges by restricting parametric flexibility, particularly resulting in less flexible dwell-time distributions, and it often necessitates comparisons between models with varying structures (i.e. different numbers of substages). Here, we propose a Geometric Chain Trick (GCT) employing a fixed number of substages with transition rates forming a geometric sequence. Through mathematical analysis, simulation, and model fitting, we demonstrate that our approach retains the principal benefits of the LCT while providing increased flexibility and facilitating easier data fitting. Compared to the recently introduced Generalized Linear Chain Trick (GLCT), which incorporates higher-dimensional auxiliary stages to model delays in transitions between main compartments, our method offers a more direct means of achieving flexibility. Although this study focuses on the application of the LCT and GCT in the epidemic SIR (susceptible-infectious-recovered) model, the potential applications of our approach span a broad array of dynamical models, including those for infectious diseases and other systems (e.g. population dynamics, biochemical reaction networks), offering advantages in computational efficiency, ease of parameter estimation, and the flexibility of time distributions.
